---
title: "Core121 Weissman 2021 part2"
output: pdf_document
---


# Data Visualizations with R workshop
## CORE 121, Professor Weissman-Unni
## September 17th, 2021


### Goals of this workshop:

* Refresh on Loading Data Sets
* Learn the basics of ggplot for visualizations
* Practice creating histograms, boxplots, line graphs, bar graphs, and pie charts
* Save a plots as an image file



#### Load Palmer Penguin data set

(Or, could use a data set better suited to class)

USE FAKE DATA SIMILAR TO THEIR DATA STRUCTURE, AND PIVOT


```{r}
#install.packages("palmerpenguins")

library(palmerpenguins)
#data(package = 'palmerpenguins')

penguins


```



#### ggplot basics


ggplot is an R package for the purpose of creating data visualizations. The "gg" translates to "grammar of graphics", and is founded in the idea that all data visualizations are comprised of three components:

* data set
* aesthetics, or visual marks that represent the data (i.e. the stuff that you see)
* geometric objects, or "geoms" (e.g the type of plot)

ggplot is part of the "tidyverse" - a series of packages that share common strategies for working with data and generating visualizations. In last Wednesday's workshop, we used some tidyverse packages (readr, dplyr) for reading and filtering datasets. Fortunately, there's a simple way to load all these libraries at once, as we'll use various functions spread throughout different packages in the tidyverse:

```{r}
library(tidyverse)
library(janitor)
```


#### Histograms

ex

```{r}

ggplot(data=penguins, aes(x=body_mass_g))+geom_histogram()

```



Show avg's in bar graph or box plot


leave line graph/2 variable stuff later



#### Line Graphs


```{r}

#ex
ggplot(data=penguins, mapping=aes(x=body_mass_g, y=flipper_length_mm))+geom_line()

```








#### Box Plots


ex
ggplot(mpg, aes(class, hwy))+ geom_boxplot()

maybe group by species?

```{r}

```




#### Bar graphs

ggplot(data, aes(x=name, y=value)) +geom_bar(stat = "identity")





#### Pie charts

# Basic piechart
ggplot(data, aes(x="", y=value, fill=group)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)




ggsave


```{r}
x <- read.csv("data.csv", header = TRUE, stringsAsFactors = FALSE)
x<-subset(x, select=-c(Notes))
#x<-read_csv("data.csv")
x

rownames(x) <- x$Variable
x$Variable <- NULL
#head(x)


y <- as.data.frame(t(as.matrix(x)))

#y<-as_tibble(t(as.matrix(x)))
#colnames(y) <- y[1]
y<-as_tibble(y)
y$Dates<-c("9/9","9/10","9/11","9/12","9/13","9/14")
y <- y %>%
  select(Dates, everything())
y$Dates <- as.Date(y$Dates , format = "%m/%d")
#colnames(y) = y[2, ]
 #y[-nrow(y),]
y
#y<-x%>%spread(key=9/9)
#y
```


```{r}
z<-read_csv("data.csv")
#z
# drop Notes
z<-subset(z, select=-c(Notes))
#z
#inprogress<-gather(z,date,value,z$`9/9`:z$`9/14`)
#head(inprogress)
zz<-z %>%
  gather(day,value,z$`9/10`:z$`9/10`) %>%
  spread(Variable,value)

zz
```



